Mechanisation of Session-based Lambda Calculus Type Systems

### 2014-11-26 proposal/proposal.tex
Expanded the section session polymorphism.

@@

Some recent mechanisation effort focuses on providing a basis for studying
linearity within type systems. This work is of interest to the current project
since most session-based type systems assume a strictly linear type system
which requires re-binding of channel identifiers, such as in the functional
setting of \citeauthor{Gay:2010:LAST}. As discussed previously, it would be
interesting to study a system for supporting side-effecting functions (aka
closures) for session channels. In order to achieve this, operational
semantics for GV would need to be defined, and the recent draft by
\citeauthor{Lindley:2014:SPS} may provide some useful insights on how best to
do this. \citeauthor{Mazurak:2010:LLT} present an extension to System F
(termed \fpop) with a notion describing types as having linear or non-linear
\textit{kind}~\cite{Mazurak:2010:LLT}. This extension is motivated by previous
work on incorporating linearity into type systems. Overall, previous attempts
either hamper the inclusion of desirable programming features (such as
polymorphism), or do not adequately reflect non-linearity as the common case
leading to awkward programming. \fpop avoids these deficiencies by
categorising types into kinds whilst maintaining similar semantics to System
F. Mechanised proofs for type soundness and parametricity are presented, and
the semantics allow non-linear types to be treated as linear. Examples show
the system can provide a wide range of permissions on interfaces, from full
exclusive access to read-only shared references. The system is very close to
the Alms language described by \citeauthor{Aldrich:2009} but has the advantage
that the soundness results (type safety and parametricity) have been entirely
mechanised in Coq. It would be interesting to see if one could encode the
``adopt'' and ``focus'' constructs in this system to compare its
expressiveness to Alms. The ``adopt'' rule can be encoded similarly to the
ShareRef example~\cite{Mazurak:2010:LLT}. It is not immediately clear how to
encode the ``focus'' rule within the present semantics, as one needs to
prevent non-linear access to its argument implying the presence of knowledge
about aliasing within the typing judgements; absence of aliasing is not
identified with linear kinded types in \fpop.

@@

\citeauthor{Goto:2014}~\cite{Goto:2014} provide a $\pi$-calculus session-based
type system providing session polymorphism. The system is more general than
session polymorphism via subtyping~\cite{Gay:2005:SST}. more general how?

The authors mechanise their type system and its properties in the Coq proof
assistant. However, the system mechanised is restricted to the $\pi$-calculus
and the main focus is proving soundness properties for the deduction rules
allowing polymorphism on input, rather than a more general basis for studying
programming with session types.

@@

>>>>

\input{macros}

@@

Some recent mechanisation effort focuses on providing a basis for studying
linearity within type systems. This work is of interest to the current project
since most session-based type systems assume a strictly linear type system
which requires re-binding of channel identifiers, such as in the functional
setting of \citeauthor{Gay:2010:LAST}. \citeauthor{Mazurak:2010:LLT} present
an extension to System F (termed \fpop) with a notion describing types as
having linear or non-linear \textit{kind}~\cite{Mazurak:2010:LLT}. This
extension is motivated by previous work on incorporating linearity into type
systems. Overall, previous attempts either hamper the inclusion of desirable
programming features (such as polymorphism), or do not adequately reflect
non-linearity as the common case leading to awkward programming. \fpop avoids
these deficiencies by categorising types into kinds whilst maintaining similar
semantics to System F. Mechanised proofs for type soundness and parametricity
are presented, and the semantics allow non-linear types to be treated as
linear. Examples show the system can provide a wide range of permissions on
interfaces, from full exclusive access to read-only shared references. The
system is very close to the Alms language described by
\citeauthor{Aldrich:2009} but has the advantage that the soundness results
(type safety and parametricity) have been entirely mechanised in Coq. It would
be interesting to see if one could encode the ``adopt'' and ``focus''
constructs in this system to compare its expressiveness to Alms. The ``adopt''
rule can be encoded similarly to the ShareRef
example~\cite{Mazurak:2010:LLT}. It is not immediately clear how to encode the
``focus'' rule within the present semantics, as one needs to prevent
non-linear access to its argument implying the presence of knowledge about
aliasing within the typing judgements; absence of aliasing is not identified
with linear kinded types in \fpop.

@@

\citeauthor{Goto:2014}~\cite{Goto:2014} provide a $\pi$-calculus session-based
type system providing session polymorphism. The system is more general than
session polymorphism via subtyping~\cite{Gay:2005:SST} in that it is capable
of typing a generic forwarding process between mutually dual session
types. Defining such a process is not possible using subtyping since one
wishes to restrict the types to be dual, not a subtype of some other
type. This more general form of polymorphism is achieved by permitting input
on all session types (including \lstinline{end}, the terminal session type)
and then employing transition hypotheses to ensure the type permits such
input. The authors mechanise their type system and its properties in the Coq
proof assistant. However, the mechanised system is restricted to the
$\pi$-calculus and the main focus is proving soundness properties for the
deduction rules allowing polymorphism, rather than a more general basis for
studying programming with session types.

@@

As discussed previously, it would be
interesting to study a system for supporting side-effecting functions (aka
closures) for session channels. In order to achieve this, operational
semantics for GV would need to be defined, and the recent draft by
\citeauthor{Lindley:2014:SPS} may provide some useful insights on how best to
do this.
